---
title: "COVID19 Mortality rate prediction. IML '20 Homework 1."
author: "Åukasz Grad"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r, include=F}
library(tidyverse)
library(DALEX)
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(caret)
library(glmnet)

theme_set(theme_economist())
```

Read and show raw data.

```{r, include=F}
covid_hubei = read_csv('data/COVID19_2020_open_line_list_Hubei.csv')
covid_outside = read_csv('data/COVID19_2020_open_line_list_outside_Hubei.csv')
```

```{r}
common_cols = intersect(colnames(covid_hubei), colnames(covid_outside))
covid = covid_hubei %>% 
    mutate(admin_id = as.character(admin_id)) %>%
    bind_rows(covid_outside) %>%
    select(common_cols)
covid %>% head
```

We have a lot of missing values and data in general is messy. The column of interest is "outcome"

```{r}
covid %>%
    summarise_all(~ mean(is.na(.)))
```

We have 59 cases of confirmed deaths.

```{r}
death_vals = c('death', 'died')
covid %>% filter(outcome %in% death_vals)
```

```{r, include=F}
clean_string <- function(string){
    temp <- tolower(string)
    temp <- stringr::str_replace_all(temp,"[^a-zA-Z\\s]", " ")
    temp <- stringr::str_replace_all(temp,"[\\s]+", " ")
    temp <- stringr::str_split(temp, " ")[[1]]
    indexes <- which(temp == "")
    if(length(indexes) > 0){
      temp <- temp[-indexes]
    } 
    return(temp)
}
```

Let's see the most common early symptoms

```{r}
symptoms_data = lapply(covid$symptoms, clean_string)
symptoms_all = Reduce(c, symptoms_data)
symptoms_all_sorted = sort(table(symptoms_all), decreasing = T)
symptoms_all_sorted[1:15]
```

Prepare clean dataset and show some plots

```{r}
frequent_countries = c('china')
frequent_symptoms = c('pneumonia|pneumonitis', 'pain', 'fever', 
                      'cough', 'sore throat', 'fatigue', 'headache', 'dyspnea')

covid_train = covid %>%
    mutate(
        outcome = ifelse(outcome %in% death_vals, 'dead', 'recovered'),
        country = ifelse(tolower(country) %in% frequent_countries, country, 'other'),
        in_wuhan = !as.logical(`wuhan(0)_not_wuhan(1)`),
        age = ifelse(str_detect(age, "[:digit:]{2}-[:digit:]{2}"),
                     0.5 * as.numeric(str_sub(age, 1, 2)) + 0.5 * as.numeric(str_sub(age, 3, 4)),
                     as.numeric(age)),
        sex = tolower(sex),
        chronic_disease = ifelse(is.na(chronic_disease_binary),
                                'NA',
                                as.character(chronic_disease_binary))
    ) %>%
    select(age, sex, country, chronic_disease, outcome, in_wuhan, symptoms) %>%
    filter(!is.na(age)) %>%
    filter(!is.na(sex))

for (symptom in frequent_symptoms) {
    covid_train[symptom] = str_detect(symptom, covid_train$symptoms)
    covid_train[symptom][is.na(covid_train[symptom])] = F
}

covid_train = covid_train %>% select(-symptoms)

covid_train %>% head
```

China has more deaths, age is heavily influencing mortality rate.

```{r}
p1 = covid_train %>%
    ggplot(aes(outcome, fill = country)) +
    geom_bar(position = 'dodge') +
    scale_fill_economist() +
    coord_flip()

p2 = covid_train %>%
    ggplot(aes(outcome, age)) +
    geom_boxplot()

grid.arrange(p1, p2, nrow = 1)
```

Chronic disease is critical for the outcome, but data may be biased. It is also heavily correlated with age.

```{r}
covid_train %>%
    mutate(chronic_disease_known = as.factor(chronic_disease == '1')) %>%
    ggplot(aes(age, as.numeric(outcome == 'dead'), colour = chronic_disease_known)) +
    geom_smooth(method = 'loess') +
    scale_color_economist() +
    ylab('Mortality rate')
```

```{r}
covid_train %>%
    ggplot(aes(chronic_disease, fill = outcome)) +
    geom_bar(position = 'dodge') +
    scale_fill_economist() +
    scale_y_log10()
```

```{r}
covid_train %>%
    ggplot(aes(age, fill = chronic_disease)) +
    geom_histogram(stat = 'density', position = 'identity', alpha = 0.5) +
    scale_fill_economist()
```

We have very little data about dead patients with early symptoms

```{r}
covid_train %>%
    summarise_at(.vars = frequent_symptoms, 
                 list(mort = ~sum((outcome == 'dead') * (.))/sum(.), sum = ~sum(.)))
```

```{r, include=F}
covid_train = covid_train %>% select(-in_wuhan)
covid_train = covid_train[sample(1:nrow(covid_train), size = nrow(covid_train), replace = F),]
```

Create simple dataset split and fit Logistic Lasso with CV

```{r}
covid_train = covid_train %>%
    mutate_if(is.character, as.factor)

train_mask = caret::createDataPartition(covid_train$outcome, p = 0.66)[[1]]
data_train = covid_train[train_mask,]
data_valid = covid_train[-train_mask,]

xlevs_train = data_train %>% select(-outcome)
xlevs <- lapply(xlevs_train[,sapply(xlevs_train, is.factor), drop = F], function(j){
    levels(j)
})
```

```{r}
pos = function(x, s, u = NULL) {
    x = ifelse(x > s, x - s, 0)
    if (!is.null(u))
        x = ifelse(x > (u - s), u - s, x)
    x
}

dtrain = model.matrix(~ age + pos(age, 50) - 1 + ., 
                      data = data_train %>% select(-outcome))
dvalid = model.matrix(~ age + pos(age, 50) - 1 + ., 
                      data = data_valid %>% select(-outcome))
cv_glmnet = cv.glmnet(dtrain, data_train$outcome == 'dead', family = 'binomial', nfolds = 10, 
                   type.measure = 'deviance', keep = TRUE)
plot(cv_glmnet)
```

Coefficients for best model

```{r}
best_ind = which(cv_glmnet$lambda == cv_glmnet$lambda.min)
beta = cv_glmnet$glmnet.fit$beta[, best_ind]
beta
```

Train 2 more GLM models with selected features

```{r}
lrm = glm(outcome == 'dead' ~ age + sex + country +
                              fever +I((chronic_disease == '1')),
          family = 'binomial', data = data_train)
lrm_simple = glm(outcome == 'dead' ~ age + sex + country + fever,
                 family = 'binomial', data = data_train)

summary(lrm)
summary(lrm_simple)
```

```{r, include=F}
pred_data = tibble(y = data_valid$outcome == 'dead', 
       pred_link = predict(cv_glmnet, dvalid, type='link', s = 'lambda.min'),
       pred = predict(cv_glmnet, dvalid, type='response', s = 'lambda.min'))
pred_data %>%
    ggplot(aes(pred, fill = y)) +
    geom_histogram(stat = 'density', alpha = 0.5) + 
    xlim(0, 1)
```

```{r, include=F}
sigmoid = function(x) 1 / (1 + exp(x))
logit = function(x) log(x / (1 - x))

th = sigmoid(1) * 100
th
```

Use DALEX library to create explainers :-) and show validation metrics - cutoff selected manually, but it's not that important. 

```{r, include=F}
th = 25

lrm_explain = DALEX::explain(lrm, data = data_valid %>% select(-outcome), 
                             y = data_valid$outcome == 'dead', 
                             type = 'classification', label = 'GLM', 
                             predict_function = function(m, d) yhat(m, d) * 100)
lrm_simple_explain = DALEX::explain(lrm_simple, data = data_valid %>% select(-outcome),
                                    y = data_valid$outcome == 'dead', 
                                    type = 'classification', label = 'Simple GLM',
                                    predict_function = function(m, d) yhat(m, d) * 100)

glmnet_predict = function(model, data) {
    data = model.matrix(~ age + pos(age, 50) - 1 + ., data = data, xlev = xlevs)
    predict(model, data, type = 'response', s = 'lambda.min')[, 1] * 100
}

cv_glmnet_explain = DALEX::explain(cv_glmnet, data = data_valid %>% select(-outcome),
                                    y = data_valid$outcome == 'dead', 
                                    type = 'classification', label = 'GLMNET',
                                    predict_function = glmnet_predict)
```

GLMNET and GLM model with chronic_disease feature have comparable fit. We will use GLM to present explanations

```{r}
lrm_perf = DALEX::model_performance(lrm_explain, cutoff = th)
lrm_simple_perf = DALEX::model_performance(lrm_simple_explain, cutoff = th)
cv_glmnet_perf = DALEX::model_performance(cv_glmnet_explain, cutoff = th)
lrm_perf
lrm_simple_perf
cv_glmnet_perf
```

```{r}
p1 <- plot(lrm_perf, lrm_simple_perf, cv_glmnet_perf, geom = "roc")
p2 <- plot(lrm_perf, lrm_simple_perf, cv_glmnet_perf, geom = "lift")

library(patchwork)
p1 + p2
```

Present BreakDown with interactions for manually created cases.

```{r, include=F}
par(mfrow = c(2, 2))

explain_data = tibble(
    age = c(5, 15, 30, 45, 50, 55, 60, 65, 70, 75, 80, 85),
    country = c('China', 'other', 'China', 'other', 
                'China', 'other', 'China', 'other', 'China',
                'other', 'China', 'other'),
    sex = c('male', 'male', 'female', 'female', 'male', 'male', 
            'female', 'female', 'male', 'male', 'female', 'female'),
    chronic_disease = c('0', '0', '1', '0', '0', '1', '0', '0', '1', '0', '1', '0'),
    fever = c(F, T, F, T, F, T, F, T, F, T, F, F)
)
dexplain = model.matrix(~ age + sex + country + pos(age, 50) - 1, 
                        data = explain_data)

lrm_exs = list()
ps = list()
for (i in 1:nrow(explain_data)) {
    lrm_exs[[i]] = DALEX::variable_attribution(lrm_explain, 
                                               new_observation = explain_data[i, ], 
                                               type = 'break_down_interactions')
    ps[[i]] = plot(lrm_exs[[i]])
    if (i == 1) {
        ps[[i]] = ps[[i]] + ggtitle('Mortality rate COVID 19')
    }
    ps[[i]] = ps[[i]] + scale_y_continuous(expand = expand_scale(c(0, 0.4)), name = NULL)
    ps[[i]]$data$right_side = pmax(ps[[i]]$data$prev, ps[[i]]$data$cumulative)
    ps[[i]]$data$pretty_text = paste(ps[[i]]$data$pretty_text, '%')
}
```

```{r}
explain_data
```

```{r, fig.height=12, fig.width=10}
grid.arrange(ps[[1]], ps[[2]], ps[[3]], ps[[4]],
             ps[[5]], ps[[6]], ps[[7]], ps[[8]],
             ps[[9]], ps[[10]], ps[[11]], ps[[12]], nrow = 4)
```
